---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r}
library(dplyr)
library(ggplot2)
```


## Loading and preprocessing the data
```{r}
if (!file.exists('activity.csv')) {
  unzip(zipfile = "activity.zip")
}

activityData <- read.csv(file="activity.csv", header=TRUE,)
```



## What is mean total number of steps taken per day?
```{r}

# Calculate the total steps taken per day
totalstepsperday <- activityData %>% group_by(date) %>% summarise(totalsteps = sum(steps))

# Make a histogram of the total number of steps taken per day
hist(totalstepsperday$totalsteps,
     main = "Total Steps per Day",
     xlab = "Number of Steps")

meansteps <- mean(totalstepsperday$totalsteps, na.rm = TRUE)
mediansteps <- median(totalstepsperday$totalsteps, na.rm = TRUE)
```
mean of the total number of steps taken per day `r meansteps`

median of the total number of steps taken per day `r mediansteps`

## What is the average daily activity pattern?
```{r}

# Make a time-series plot of the 5-minute interval and the average number of
# steps taken, averaged acoss all days.
avgstepsperinterval <- activityData %>% group_by(interval) %>% summarise( meansteps = mean(steps, na.rm = TRUE))

ggplot(data = avgstepsperinterval, aes(x = interval, y = meansteps)) +
  geom_line() +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))

# Which 5-minute interval across all days contain the maximum number of steps
maxsteps <- avgstepsperinterval[which.max(avgstepsperinterval$meansteps),]
```

The 5-minute interval that contains the maximun number of steps on average across all the days in the dataset is `r maxsteps[,1]`

## Imputing missing values
```{r}
totalNAs <- sum(nrow(is.na(activityData)))
```
The total number of missing values is `r totalNAs`. I will replace these missing values with the 5-day average of that respective interval.


```{r}
# Create a new dataset that is equal to the original dataset but with 
# the missing data filled in.
imp_activityData <- transform(activityData,
                              steps = ifelse(is.na(activityData$steps),                                          avgstepsperinterval$meansteps[match(activityData$interval, avgstepsperinterval$interval)],
                                             activityData$steps))

# Make a histogram of the total number of steps taken each day and
# and report the mean and median.
impStepsByInt <- aggregate(steps ~ date, imp_activityData, FUN=sum)
hist(impStepsByInt$steps,
     main = "Imputed Number of Steps Per Day",
     xlab = "Number of Steps")
```

```{r}

impMeanSteps <- mean(impStepsByInt$steps, na.rm = TRUE)
impMedSteps <- median(impStepsByInt$steps)
diffMean = impMeanSteps - meansteps
diffMed = impMedSteps - mediansteps
diffTotal = sum(impStepsByInt$steps) - sum(totalstepsperday$totalsteps, na.rm = TRUE)
```

There is a difference of `r diffMean` in the mean steps of the two dataset. There is a difference of `r diffMed` in the median steps of the two dataset. There is a difference of `r diffTotal` in the total steps of the two dataset.



## Are there differences in activity patterns between weekdays and weekends

```{r}
DayType <- function(date) {
  day <- weekdays(date)
  if (day %in% c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'))
      return ("weekeday")
  else if (day %in% c('Saturday', 'Sunday'))
      return ("weekend")
  else
      stop ("Invalid Date Format.")
}
imp_activityData$date <- as.Date(imp_activityData$date)
imp_activityData$day <- sapply(imp_activityData$date, FUN = DayType)

# Make a panel plot containnig a time-series plot of the 5-minute interval
# and the average number of steps taken across all weekdays or weekends
meanStepsByDay <- aggregate(steps ~ interval + day, imp_activityData, mean)
ggplot(data = meanStepsByDay, aes(x = interval, y = steps)) + 
  geom_line() +
  facet_grid(day ~ .) +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```

